Towards a wide-coverage grammar for Swedish

### 2011-12-19 doc/report/report.tex
try to parse nps

The Swedish version of the Core Language Engine (CLE) \cite{gamback} % \cite{cle}
gives a full syntactic analysis as well as semantics represented in xx. A
translation for conversations?? to English was implemented. The work was further developed in the spoken
language blabla (xx,xx). Unfortunately, it is no longer available. The coverage of the Swedish
CLE is also reported to be very limited (Nivre,2002).\\

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar. When it comes to parsing,
however, we do not get far without robustness.
The grammar in itself is by no means robust, and just one unknown word,
unexpected punctuation
or missing phrase will cause the parsing of the whole sentence to fail. 
Our goal is to parse Talbanken, but considering that no syntactic theory is capable
of covering a whole natural language, our grammar when used without a
robust layer should just be seen as an important
fragment.\\

@@

By the renewed import of Saldo, we doubled the size of the lexicon and thereby
added many of the commonly used irregular words that were missing from the older
version. This is of course a big improvement.
However, the lexical part still requires much work before it can be made use of.
We need valency information to make a good analysis. The lexicon is also too
big to use in a trivial manner, its size makes the incremental parsing algorithm
use too much memory. To solve this, we need to use the lexicon data more cleverly.

@@

\section{Future Work}
\label{sec:future}
Although this one part of the project is finished, there are many interesting future
directions. As a natural language processing-project ~ it will hardely ever be entirely
complete.
The future work described in this section is divided into two categories:
the ones aiming at making the parser robust and the ones that can be
seen  as extensions of the work done so far.

\subsection{Enhancements and complements}
\subsubsection{Grammar}
All imaginable grammatical constructs cannot be cover by our grammar.
Some work must be left for the robustness, but some specific constructions
are desirable and interesting to implement. A few examples are listed here:\\

In many Germanic? languages \textit{object shifts} are common.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
Personal pronouns are allowed to shift position with the negation in main clauses.
(read).
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & *Vi & ser&  varandra&  inte&~~ & b.& * Jag &ser &huset &inte\\
   & \emph{~we} & \emph{see}&  \emph{each other}&  \emph{not} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
\end{tabular}}

This is common in relaxed? language, but since it is rarely found in
Talbanken's part P, which we have been using, it has not been highly prioritised.\\

Sentences like
'hon är inte dum, hon.' 'hon är inte dum, flickan' 'min syster, hon är inte dum, hon'
'min syster är inte dum, hon'
We have a clause of which we need to know the natural gender and number of the subject, since it
must agree with the added pronoun. sag s 439. fullständigt outforskade!

%or just implement this and remove it here?
The question is how many types we should give dependent types. CN could also need it,
in order to say
'han mäter avståndet från sitt hus till sitt jobb' but now `avståndet från sitt hus är stort'.
The analysis of this sentence is that
avståndet is a noun of type \verb-N3-, which chooses the prepositions ``från" and ``till"
and needs to noun phrases to form a \verb-CN- -- ``avståndet från sitt hus till".
Now also \verb-CN- needs to carry about whether it is possible to use it as a subject.

Nouns used without any explicit definitness can at this stage be parsed as mass
nouns.
\enumsentence{\shortexnt{3}
{Jag & dricker& vatten}
{\emph{I}&\emph{drink}&\emph{water} }
}
This is not the correct analyse for `hund' in sentence \ref{sent:hund}.
\enumsentence{\shortex{4}
{Vi & ska& köpa & hund}
{\emph{we} &\emph{will}& \emph{buy} & \emph{dog} }
{We are going to buy a dog}}\label{sent:hund}
This sentence is instead an example of \textit{bare indefinites}, used for
... what?\\

\verb| X : NP x + CN + NP --better|\\
\verb| X : CN x + CN + CN|
check apposCN.
\textbf{måttsattribut}
\enumsentence{
{Antalet bilar, ett antal bilar, ett stort antal}}
Genitive; Not like 'sonen till mig' -> antalet bilars??\\

\subsubsection{Lexicon}
Our lexicon still lacks some part that can be imported from Saldo.
The multipleword verbs (\verb-vbm-) and multiple word nouns (\verb-nm-) should
be imported to an idiom lexicon, which can be used as extensions to the main
lexicon.
For the words that we tried but failed to import, a tool for lexicon
acquisition could be used. The tool developed in the previous part of
this project\footnote{??} would be suitable.
All in all, it should be ensured that we use as much information as
possible from our resources.
%The list of not-imported can be used for adding other kinds of lexicons - idioms, have pos-tags.
%Also to analyse what's wrong, or use the words with a word guesser. The code
%has been adapted for this.
%For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
%category. How this works, look at tags. Small amount, can be done manually.

\subsection{Making the grammar robust}
To achieve our goal of free parsing, we need to use statistical methods.
The parser needs to be able to handle
ellipses and long sentences, names and numbers,
compounds words and unknown constructions.
\subsubsection{Parallel parsing/Chunk parsing}
Instead of parsing a whole sentence in one go, we would like to
parallely parse parts of it and then try to combine the pieces.
This method should be faster then traditional parsing and would
at the same time give robustness.
Components that cannot be combined by any grammar rule are combined by a 
meta function. We might end up with partial parse trees, but we would
be able to parse at least parts of every sentence and give a more
interesting result to the user.

@@

We need to come up with better methods for using the lexicon when
parsing, since its time and memory consumption is very high.
One possibility is of course to do 
deeper refinements of the parsing source code, which would be a substantial 
project. Other solutions are to to either
adapt the lexicon for its domain and hence making it smaller and faster
or to adapt the input sentences to a smaller lexicon, by preprocessing .

Further, the lexicon needs valencies since this information lays the
foundation for the GF analysis. We have already analysed how Talbanken
annotates valency information, and by using the rules made for the mapping
described in section \ref{sec:Mapping}, we could extract data from Talbanken,
showing how words are normally used. There should also be algorithms for
automatically adding the information to the lexicons. 
Many Swedish verbs can be used with different number of arguments. Adding
all this information to the lexicon does not seem to be a good idea considering
the ambiguities it will lead to and the already high time usage.
The task should instead be left to the robust layer of the parser, possibly
implemented by using external resources.

@@

\subsubsection{Probabilities}
\label{sec:futureProbabilities}
The grammar will always contain ambiguities but before returing a parse result
to the user, some analysis should be done to find the most likely tree.
When the size of the lexicon
increases, so will the ambiguities as the number of word forms overlapping each other
gets higher. Our new grammar also have many more rules which also will contribute
to increase the number to interpret each sentence.

GF already allows the lexicon to have probabilities, and by using our extracted
GF treebank, we can get reliable statisics for our functions. 
By implementing dependency probabilities, our results would be even better.

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth and coverage. for now depth,
>> robustness in in future work leads to coverage.
>> reusablilty, freely available.
>> have experiments of how to extract info from talbanken.\\
We have developed the main components
for a deep Swedish parser. %To abstract:  by combining resources.
Our main resource is the grammar formalism GF,
which allows us to start from a well-defined system for describing grammar.
The parser is derived from the grammar and will therefore validate that
a given sentence is grammatically correct according to the rules defined.
All parts, including the underlying grammar, are open-source and 
may thus be used in applications dealing with controlled natural languages.

>>>>

The Swedish version of the Core Language Engine (CLE) \cite{gamback} % \cite{cle}
gives a full syntactic analysis as well as semantics represented in xx. A
translation for conversations to English was implemented. The work was further developed in the spoken
language(xx,xx). Unfortunately, it is no longer available. The coverage of the Swedish
CLE is also reported to be very limited (Nivre,2002).\\

@@

The grammar has been extended and enhanced, and its current status is
a specialized extension of the resource grammar. When it comes to parsing,
however, we do not get far without robustness.
The grammar in itself is by no means robust, and just one 
unexpected punctuation mark, unknown word
or missing phrase will cause the parsing of the whole sentence to fail. 
Our goal is to parse Talbanken, but considering that no syntactic theory is capable
of covering a whole natural language, our grammar when used without a
robust layer should just be seen as an important
fragment.\\

@@

By the renewed import of Saldo, we have doubled the size of the lexicon and thereby
added many of the commonly used words that were missing from the older
version. This is of course a big improvement.
However, the lexical part still requires much work before it can be made use of.
We need valency information to make a good analysis. The lexicon is also too
big to use trivially with the current techniques. Its size causes the
incremental parsing algorithm use more heap memory than what normal computers
allow. To solve this, we need to use the lexicon data more cleverly.

@@

\section{Future Work}
\label{sec:future}
Although this one part of the project is finished, there are many interesting future
directions. As a natural language processing-project it will probably never be
entirely complete.
The future work described in this section is divided into two categories:
the ones aiming at making the parser robust and the ones that can be
seen as extensions of the work done so far.

\subsection{Enhancements and complements}
\subsubsection{Grammar}
All imaginable grammatical constructs cannot be cover by our grammar.
Some work must be left for the robustness, but there are specific constructions
that are desirable and interesting to implement. A few examples are listed here:\\

\begin{itemize}
\item
\textit{Pronomial object shifts} are common in Swedish.
\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & Jag & ser&  honom& \textbf{inte}&~~ & b.&Jag &ser &\textbf{inte} &honom\\
   & \emph{I} & \emph{see}&\emph{him}&\emph{not}&&& \emph{I} &\emph{see}&\emph{not}&\emph{him}\\
\end{tabular}}
Personal pronouns are allowed to shift position with the negation in main clauses, without
the modal verbs.

\enumsentence{\begin{tabular}[t]{@{}*{11}{l@{\ }}}
a. & *Vi & ser&  varandra&  inte&~~ & b.& * Jag &ser &huset &inte\\
   & \emph{~we} & \emph{see}&  \emph{each other}&  \emph{not} && b.& \emph{~I}&\emph{see} &\emph{the house} &\emph{not}\\
\end{tabular}}

Although object shifts are frequently used, they are rarely found in
Talbanken's part P, which have been the inspiration for this project.
Therefore, it has so far not been prioritised.\\

%Sentences like
%'hon är inte dum, hon.' 'hon är inte dum, flickan' 'min syster, hon är inte dum, hon'
%'min syster är inte dum, hon'
%We have a clause of which we need to know the natural gender and number of the subject, since it
%must agree with the added pronoun. sag s 439. fullständigt outforskade!

%or just implement this and remove it here?
%The question is how many types we should give dependent types. CN could also need it,
%in order to say
%'han mäter avståndet från sitt hus till sitt jobb' but now `avståndet från sitt hus är stort'.
%The analysis of this sentence is that
%avståndet is a noun of type \verb-N3-, which chooses the prepositions ``från" and ``till"
%and needs to noun phrases to form a \verb-CN- -- ``avståndet från sitt hus till".
%Now also \verb-CN- needs to carry about whether it is possible to use it as a subject.

\item
\textit{Bare indefinites} are at this stage always parsed as mass nouns.
\enumsentence{\shortexnt{3}
{Jag & dricker& vatten}
{\emph{I}&\emph{drink}&\emph{water} }
}
This however is not the correct analyse for `hund' in sentence \ref{sent:hund}.
\enumsentence{\shortex{4}
{Vi & ska& köpa & hund}
{\emph{we} &\emph{will}& \emph{buy} & \emph{dog} }
{We are going to buy a dog}}\label{sent:hund}

\item
The implementation of reflexive pronouns can be improved.  %both code-wise
It should for example be possible to differenciate between 
object and subject control verbs.

\eenumsentence{
\item[a]\shortex{7}
{Han & ber & henne & att & städa & sitt & rum}
{\emph{he} & \emph{begs} & \emph{her} & \emph{to} & \emph{clean} & {\sc self's} & \emph{room}}
{He asks her clean her room}
\item[b]\shortex{7}
{Han& lovar& henne& att& städa &sitt & rum}
{\emph{he}& \emph{promises}& \emph{her}& \emph{to}&  \emph{clean} &\sc self's&\emph{room}}
  {He promises her to clean his room}}

%
%\verb| X : NP x + CN + NP --better|\\
%\verb| X : CN x + CN + CN|
%check apposCN.
%\textbf{måttsattribut}
%\enumsentence{
%{Antalet bilar, ett antal bilar, ett stort antal}}
%Genitive; Not like 'sonen till mig' -> antalet bilars??\\

\end{itemize}
\subsubsection{Lexicon}
Our lexicon still lacks some part that can be imported from Saldo.
The multipleword verbs (\verb-vbm-) and multiple word nouns (\verb-nm-) should
be imported to an idiom lexicon, which can be used as extensions to the main
lexicon.
For the words that we tried but failed to import, a tool for lexicon
acquisition could be used. The tool developed in the previous part of
this project\footnote{??} would be suitable.
All in all, it should be ensured that we use as much information as
possible from our resources.
%The list of not-imported can be used for adding other kinds of lexicons - idioms, have pos-tags.
%Also to analyse what's wrong, or use the words with a word guesser. The code
%has been adapted for this.
%For exapmle pronouns - dets, have combined information from Talbanken to add them in the correct
%category. How this works, look at tags. Small amount, can be done manually.

\subsection{Making the grammar robust}
To achieve our goal of free parsing, we need to use statistical methods.
The parser needs to be able to handle
ellipses and long sentences, names and numbers,
compound words and unknown constructions.
\subsubsection{Parallel parsing/Chunk parsing}
Instead of parsing a whole sentence in one go, we would like to
parallely parse parts of it and then try to combine the pieces.
This method should be faster then normal parsing and would
at the same time give robustness.
Components that cannot be combined by any grammar rule are combined by a 
meta function. We might end up with partial parse trees, but we would
be able to parse at least parts of every sentence and give a more
interesting result to the user.

@@

As noted, we additionally need to come up with better methods for using the lexicon when
parsing, since its time and memory consumption is very high.
One possibility is of course to do 
deeper refinements of the parsing source code, which would be a substantial 
project. Other solutions are to to either
adapt the automatically lexicon for its domain and hence making it smaller and
faster, or to adapt the input sentences to a smaller lexicon, by preprocessing.

Further, the lexicon needs valencies since this information lays the
foundation for the GF analysis. We have already analysed how Talbanken
annotates valency information, and by using the rules made for the mapping
described in section \ref{sec:Mapping}, we could extract data from Talbanken,
showing how words are normally used. There should also be algorithms for
automatically adding the information to the lexicon. \\
Many Swedish verbs can be used with different number of arguments. Adding
all this information to the lexicon does not seem to be a good idea considering
the ambiguities it will lead to and the already high time usage.
The task should instead be left to the robust layer of the parser, possibly
implemented by using external resources.

@@

\subsubsection{Probabilities}
\label{sec:futureProbabilities}
The grammar will always contain ambiguities but before returing the result
to the user, some analysis should be done to find the most probable tree.
When the size of the lexicon
increases, so will the ambiguities as the number of word forms overlapping each other
gets higher. Our new grammar also have many more rules which also will contribute
to increase the number to interpret each sentence.

GF already allows the lexicon to have probabilities, and by using our extracted
GF treebank, we can get more reliable statisics for our functions. 
By implementing dependency probabilities, our results would be even better.

@@

\section{Conclusion}
>> why is this significant, other grammars can use this Swedish grammar.
>> well known (comp. ling) trade-off between depth and coverage. for now depth,
>> robustness in in future work leads to coverage.
>> reusablilty, freely available.
>> have experiments of how to extract info from talbanken.\\
We have developed the main components
for a deep Swedish parser. %To abstract:  by combining resources.
Our main resource is the grammar formalism GF,
which allows us to start from a well-defined system for describing grammar.
The parser is derived from the grammar and will therefore validate that
a given sentence is grammatically correct according to the rules defined.
All parts, including the underlying grammar, are open-source and 
may thus be used in other applications.
